---
title: "Metadata metrics for cancer corpus"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Semantic metrics for cancer corpus}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE}
suppressPackageStartupMessages({
library(ggplot2)
library(plotly)
library(metametrics)
library(ssrch)
})
```

# Basic observations on a corpus of human RNA-seq studies in cancer

Using the Omicidx system, we harvested metadata about human samples
for which RNA-seq data was deposited in NCBI SRA.

We work with a subset of 1009 studies for which a cancer-related
term was present in study title as recorded at NCBI SRA.

```{r lk1}
library(ggplot2)
library(plotly)
library(metametrics)
data(study_publ_dates) # harvesting omicidx early 2019
library(lubridate)
ds_ca = DocSet_ca1009()
ds_ca
```

We accumulate (over dates of study submissions)
the set of fields used in the sample annotation of the 1009 cancer studies.
```{r lk2,cache=TRUE,echo=FALSE}
study_publ_dates = na.omit(study_publ_dates)
studs1009 = ls(docs2kw(ds_ca))  # in cancer corpus
stud_dates = as_datetime(study_publ_dates[,2])
names(stud_dates) = study_publ_dates[,1]
stud_dates = stud_dates[studs1009]  # limit to corpus
stud_dates = sort(stud_dates)
ofields = lapply(names(stud_dates), 
    function(x) names(retrieve_doc(x, ds_ca)))
freqs = table(unlist(ofields))
#sort(freqs,decreasing=TRUE)[1:20]
cumfields = ofields
for (i in 2:length(cumfields)) cumfields[[i]] = 
    union(cumfields[[i]], cumfields[[i-1]])
csiz = sapply(cumfields,length)
bag_fields_ca1009 = unique(unlist(cumfields))
nfields = length(bag_fields_ca1009)
mydf = data.frame(date_published=stud_dates, nfields=csiz)
```

The growth in size of the set of fields in use over time is displayed here:

```{r lk3}
ggplot(mydf, aes(x=date_published, y=nfields)) + geom_point()
```

```{r lkdi,echo=FALSE}
library(plotly)
ddf = data.frame(date=stud_dates[-1], newly_introduced_fields=diff(csiz),
    study=paste0(names(stud_dates[-1]), "\na"))
```

The next display is interactive -- hover over points to see study
accession number and newly introduced field names.

```{r ddd,echo=FALSE,fig.width=6}
incrs = lapply(2:length(cumfields), function(x) setdiff(cumfields[[x]],
   cumfields[[x-1]]))
incrs = unlist(lapply(incrs, function(x) paste0(x, collapse="\n")))
sn = names(stud_dates[-1])
incrs = paste(sn, incrs, sep="\n")
dddf = cbind(ddf, incrs)
g2 = ggplot(dddf, aes(x=date, y=newly_introduced_fields, text=incrs)) + geom_point()
ggplotly(g2)
```

# Reference resources for reducing metadata isolation and variability

Use of common data elements is promoted by various initiatives.
Dictionaries, thesauri, and ontologies are all relevant.  We have
examples of each in the metametrics package.

A snapshot of the Genomic Data Commons gdcdictionary, with fields
and values related to diagnosis and sample characteristics is
provided in `gdc_dx_sam`.
```{r lkref}
gdc_dx_sam
```

A table with all entries from several ontologies and the NCI Thesaurus
is provided by `load_ontolookup`:
```{r lkr2}
olook = load_ontolookup()
olook
```

## Statistics on field use

We use robust linear modeling to estimate growth in
vocabulary of fields employed over time.  The data.frame
`mydf` includes a variable `nfields` taking a value
for each study publication date.  The value of `nfields` associated
with date $d$ records the
the number of fields used to annotate all studies up
to date $d$.

```{r lknf}
library(MASS)
nsecpy = 3600*24*365
summary( mm <- rlm(nfields~I(as.numeric(date_published)/nsecpy), data=mydf))
plot(nfields~I(as.numeric(date_published)/nsecpy), data=mydf)
abline(mm)
```

# Proximity of terms in use to endorsed terminologies
