---
title: "Semantic metrics for cancer corpus"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Semantic metrics for cancer corpus}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

All based on Sean Davis' [Omicidx](https://api-omicidx.cancerdatasci.org/sra/1.0/ui/)

Basic ideas: does the set of fields in use in sample.attributes
stabilize over time?  How do new fields emerge over time?

We work with a subset of 1009 studies manifestly addressing cancer
based on parsing study titles.

```{r lk1}
library(metametrics)
data(study_publ_dates) # harvesting omicidx early 2019
library(lubridate)
load(DocSet_ca1009()) # produces ds_can1009b
assign("zipf", csvs_ca1009(), envir=environment(ds_can1009b@doc_retriever))
allstud = ls(docs2kw(ds_can1009b)) 
```

Here we assemble the cumulative (over time)
word-bag of fields used in the sample annotation of the 1009 cancer studies.
```{r lk2,cache=TRUE}
study_publ_dates = na.omit(study_publ_dates)
studs1009 = ls(docs2kw(ds_can1009b))  # in cancer corpus
stud_dates = as_datetime(study_publ_dates[,2])
names(stud_dates) = study_publ_dates[,1]
stud_dates = stud_dates[studs1009]  # limit to corpus
stud_dates = sort(stud_dates)
ofields = lapply(names(stud_dates), 
    function(x) names(retrieve_doc(x, ds_can1009b)))
cumfields = ofields
for (i in 2:length(cumfields)) cumfields[[i]] = 
    union(cumfields[[i]], cumfields[[i-1]])
csiz = sapply(cumfields,length)
```

The growth in word bag size over time:
```{r lk3}
plot(stud_dates, csiz)
```

```{r lkdi}
library(ggplot2)
library(plotly)
ddf = data.frame(date=stud_dates[-1], bag_incr=diff(csiz),
    study=paste0(names(stud_dates[-1]), "\na"))
g1 = ggplot(ddf, aes(x=date, y=bag_incr, text=study)) + geom_point()
```

An interactive plot -- hover over points to get study
accession number and "as-yet-unseen" field names.

```{r ddd}
incrs = lapply(2:length(cumfields), function(x) setdiff(cumfields[[x]],
   cumfields[[x-1]]))
incrs = unlist(lapply(incrs, function(x) paste0(x, collapse="\n")))
sn = names(stud_dates[-1])
incrs = paste(sn, incrs, sep="\n")
dddf = cbind(ddf, incrs)
g2 = ggplot(dddf, aes(x=date, y=bag_incr, text=incrs)) + geom_point()
ggplotly(g2)
```
